Optimizing Abstract Abstract Machines

### 2012-11-12 paper/paper.tex
Removed repetition in baseline.
Start of evaluation section.  Some related work.

The \emph{abstracting abstract machines} (AAM)
approach~\cite{dvanhorn:VanHorn2011Abstracting,dvanhorn:VanHorn2012Systematic}
to deriving program analyses provides a straightforward way of
transforming a programming language semantics in the form of an
abstract machine, into a family of abstract interpreters parameterized
over policies for regulating the analytic precision of the control,
environment, store, and base value domains.  By taking a
machine-oriented view of computation, it becomes possible to design,
verify, and implement program analyzers for realistic language
features typically considered difficult to model.  The approach was
originally applied to features such as higher-order functions,
stack-inspection, exceptions, laziness, first-class continuations, and
garbage collection.  It has since been used to verify actor-
\cite{local:DOsualdo:12A} and
thread-based~\cite{dvanhorn:Might2011Family} parallelism and
behavioral contracts~\cite{dvanhorn:TobinHochstadt2012Higherorder}; it
has been used to model Coq~\cite{local:harvard},
Dalvik~\cite{local:dalvik}, Erlang~\cite{local:DOsualdo:12B},
JavaScript~\cite{local:DBLP:journals/corr/abs-1109-4467}, and
Racket~\cite{dvanhorn:TobinHochstadt2012Higherorder}.

@@

@@

\cite{dvanhorn:Earl2012Introspective}

\cite{dvanhorn:wright-jagannathan-toplas98}

Other benchmarks

@@

>>>>

The \emph{abstracting abstract machines} (AAM)
approach~\cite{dvanhorn:VanHorn2011Abstracting,dvanhorn:VanHorn2012Systematic}
to deriving program analyses provides a straightforward way of
transforming a programming language semantics in the form of an
abstract machine, into a family of abstract interpreters parameterized
over policies for regulating the analytic precision of the control,
environment, store, and base value domains.  By taking a
machine-oriented view of computation, it becomes possible to design,
verify, and implement program analyzers for realistic language
features typically considered difficult to model.  The approach was
originally applied to features such as higher-order functions,
stack inspection, exceptions, laziness, first-class continuations, and
garbage collection.  It has since been used to verify actor-
\cite{local:DOsualdo:12A} and
thread-based~\cite{dvanhorn:Might2011Family} parallelism and
behavioral contracts~\cite{dvanhorn:TobinHochstadt2012Higherorder}; it
has been used to model Coq~\cite{local:harvard},
Dalvik~\cite{local:dalvik}, Erlang~\cite{local:DOsualdo:12B},
JavaScript~\cite{local:DBLP:journals/corr/abs-1109-4467}, and
Racket~\cite{dvanhorn:TobinHochstadt2012Higherorder}.

@@

We have implemented, optimized, and evaluated \emph{all} of the
machines presented in their JFP paper
\cite{dvanhorn:VanHorn2012Systematic}, with the exception of the
machine for performing pushdown flow analysis.  This includes machines
supporting higher-order functions, state, first-class control,
laziness, and stack inspection.

For each benchmark, we collect analysis times, peak memory usage, and
the rate of states-per-second explored by the analysis for each of the
optimizations discussed in section~\ref{sec:opt}, culmulatively
applied.  The analysis is stopped after consuming 30 minutes of time
or 1 gigabyte of space.  When presenting \emph{relative} numbers, we
use the timeout limits as a lower bound on the actual time required,
thus giving a conservative estimate of improvements.

All benchmarks are calculated as an average of N runs on a [MACHINE
  SPECS].

\begin{figure}
\begin{tiny}
\begin{alltt}
Program  |  LOC | B. Time | O. Time | B. Space | O. Space | B. Rate | O. Rate
----------------------------------------------------------------------------
nucleic  | 3508 |
matrix   |  747 |
nbody    | 1434 |
earley   |  667 |
maze     |  681 |
church   |   41 |
lattice  |  214 |
boyer    |  642 |
mbrotz   |   68 |
\end{alltt}
\end{tiny}
\end{figure}

\begin{itemize}

\item {\bf nucleic}: a floating-point intensive application taken from
  molecular biology that has been used widely in benchmarking
  functional language
  implementations~\cite{dvanhorn:Hartel1996Benchmarking} and analyses
  (e.g.~\cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98}).
  It is a constraint satisfaction algorithm used to determine the
  three-dimensional structure of nucleic acids.

\item {\bf matrix} tests whether M is maximal among all matrices of
  the same dimension obtainable by simple reordering of rows and
  columns and negation of any subset of rows and columns.  It is
  written in continuation-passing style (used in
  \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98}).

\item {\bf nbody}: implements the Greengard multipole algorithm for
  computing gravitational forces on point masses distributed uniformly
  in a cube (used in
  \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98}).

\item {\bf earley}:  Earley's parser written by Marc Feeley.

\item {\bf maze}: generates a random maze and finds a path (used in
  \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98})
  Makes use of Scheme's {\tt call-with-current-continuation}
  operation.

\item {\bf church}: tests distributivity of multiplication over
  addition for a few Church numerals (introduced by
  \cite{dvanhorn:Vardoulakis2011CFA2}).

\item {\bf lattice}: enumerates the maps between two lattices (used in
  \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98})

\item {\bf boyer}: a term-rewriting theorem prover (used in
  \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98})

\item {\bf mbrotZ}: generates Mandelbrot set fractal using complex
  numbers.

\item {\bf graphs}: counts the number of directed graphs with a
  distinguished root and \(k\) vertices, each having out-degree at
  most 2. It makes extensive use of higher-order procedures and is
  written in a continuation-passing style (used by
  \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98})

\end{itemize}

@@

@@

\paragraph{Abstracting Abstract Machines}

This work clearly closely follows Van Horn and Might's original papers
on abstracting abstract
machines~\cite{dvanhorn:VanHorn2011Abstracting,dvanhorn:VanHorn2012Systematic},
which in turn is one piece of the large body of research on flow
analysis for higher-order languages (see
Midtgaard~\cite{dvanhorn:Midtgaard2011Controlflow} for a thorough
survey).  The AAM approach sits at the confluence of two major lines
of research: (1) the study of abstract
machines~\cite{dvanhorn:landin-64} and their systematic
construction~\cite{dvanhorn:reynolds-hosc98}, and (2) the theory of
abstract interpretation
\cite{dvanhorn:Cousot:1977:AI,dvanhorn:Cousot:1979:Galois}.

\paragraph{Frameworks for flow analysis of higher-order programs}

The analysis most similar to that presented in section~\ref{sec:aam}
is the infinitary control-flow analysis of Nielson and
Nielson~\cite{dvanhorn:nielson-nielson-popl97} and the unified
treatment of flow analysis by Jagannathan and
Weeks~\cite{dvanhorn:jagannathan-weeks-popl95}.  Both are
parameterized in such a way that in the limit, the analysis is
equivalent to an interreter for the language, just as is the case
here.  What is different is that both give a constraint-based
formulation of the abstract semantics rather than a finite machine
model.

\paragraph{Abstract compilation}
